---
title: "Project 2 Codebook"
author: "Yingxi Kong"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load necessary packages
library(tidyverse)
library(mice)
library(gt)
library(gtsummary)
library(kableExtra)
library(RColorBrewer)
library(scico)
library(caret)
library(glmnet)
library(pROC)
library(predtools)
library(gridExtra)
library(ggpubr)
library(patchwork)
```


```{r}
# set working directory
# Windows
# setwd("C:/Users/yingx/OneDrive/Desktop/Fall 2024/PHP 2550/Data/")

# Mac
setwd("~/Desktop/Fall 2024/PHP 2550/Data/")

# read in data
data <- read.csv("project2.csv")

data[, c("abst", "Var", "BA", "sex_ps", "NHW", 
         "Black", "Hisp", "inc", "edu", "ftcd.5.mins", 
         "otherdiag", "antidepmed", "mde_curr", 
         "Only.Menthol")] <- lapply(data[, c("abst", "Var", "BA", "sex_ps", "NHW", 
                                             "Black", "Hisp", "inc", "edu", 
                                             "ftcd.5.mins", "otherdiag", "antidepmed", 
                                             "mde_curr", "Only.Menthol")], as.factor)
```

`Table 1` presents an overall summary statistics of patients' baseline characteristics by their behavioral and pharmacological treatment assignment. Since our study is a $2 \times 2$, factorial, randomized, placebo-controlled trial, patients are randomly assigned to either behavioral activation for smoking cessation group (BASC) or standard behavioral treatment group (ST) and either varenicline or placebo blister pack. Patients can be categorized into four treatment arm groups: BASC + placebo, BASC + varenicline, ST + placebo, and ST + varenicline. Seeing from `Table 1`, the two placebo groups both have 68 obervations while the two varenicline groups both have 83 observations. 

Most variables are evenly distributed across the four treatment arms, which reflects successful randomization in this factorial trial. However, a few key factors, such as socioeconomic indicators (income and education) and specific mental health variables (MDD status, DSM-5 diagnoses), exhibit slight variations that may influence outcomes. Notably, treatment arms with varenicline show higher abstinence rates than placebo groups, suggesting the potential efficacy of this pharmacotherapy in combination with behavioral interventions. While many baseline characteristics are evenly distributed across groups, some may still function as moderators, potentially interacting with treatment assignment to affect abstinence success. 

```{r}
# Recode factor levels in the dataset
averaged_data_factor <- data %>%
  mutate(abst = fct_recode(as.factor(abst), "Yes" = "1", "No" = "0"),
         inc = fct_recode(as.factor(inc), 
                          "Less than $20,000" = "1", 
                          "$20,000-35,000" = "2", 
                          "$35,001-50,000" = "3", 
                          "$50,001-75,000" = "4", 
                          "More than $75,000" = "5"),
         sex_ps = fct_recode(as.factor(sex_ps), "Male" = "1", "Female" = "2"),
         edu = fct_recode(as.factor(edu), 
                          "Grade School" = "1", 
                          "Some high school" = "2", 
                          "High school graduate or GED" = "3", 
                          "Some college/technical school" = "4", 
                          "College graduate" = "5"),
         ftcd.5.mins = fct_recode(as.factor(ftcd.5.mins), "Yes" = "1", "No" = "0"),
         otherdiag = fct_recode(as.factor(otherdiag), "Yes" = "1", "No" = "0"),
         antidepmed = fct_recode(as.factor(antidepmed), "Yes" = "1", "No" = "0"),
         mde_curr = fct_recode(as.factor(mde_curr), "Current MDD" = "1", "Past MDD" = "0"),
         Only.Menthol = fct_recode(as.factor(Only.Menthol), "Yes" = "1", "No" = "0"),
         race = case_when(Black == 1 ~ "Black",
                          Hisp == 1 ~ "Hispanic",
                          NHW == 1 ~ "Non-Hispanic White",
                          TRUE ~ "Other"),
         trt = case_when(Var == 1 & BA == 1 ~ "BASC + varenicline",
                         Var == 0 & BA == 1 ~ "BASC + placebo",
                         Var == 1 & BA == 0 ~ "ST + varenicline",
                         Var == 0 & BA == 0 ~ "ST + placebo",
                         TRUE ~ NA_character_))

averaged_data_factor <- averaged_data_factor %>%
  mutate(inc = fct_relevel(inc, "Less than $20,000", "$20,000-35,000", 
                           "$35,001-50,000", "$50,001-75,000", "More than $75,000"),
         edu = fct_relevel(edu, "Grade School", "Some high school", "High school graduate or GED",
                           "Some college/technical school", "College graduate"))

# Now create the summary table
summary_table <- averaged_data_factor %>%
  select(-c("id", "Var", "BA", "Black", "Hisp", "NHW")) %>%
  tbl_summary(by = trt, label = list(abst ~ "Smoking abstinence",
                                     race ~ "Race",
                                     age_ps ~ "Age",
                                     sex_ps ~ "Sex",
                                     inc ~ "Income",
                                     edu ~ "Education",
                                     ftcd_score ~ "FTCD score",
                                     ftcd.5.mins ~ "Smoking within 5 mins of waking up",
                                     bdi_score_w00 ~ "BDI score",
                                     cpd_ps ~ "Cigarettes smoked per day",
                                     crv_total_pq1 ~ "Cigarette reward value",
                                     hedonsum_n_pq1 ~ "Pleasurable events (substitute reinforcers)",
                                     hedonsum_y_pq1 ~ "Pleasurable events (complementary reinforcers)",
                                     shaps_score_pq1 ~ "Anhedonia",
                                     otherdiag ~ "Other lifetime DSM-5 diagnosis",
                                     antidepmed ~ "Taking antidepressant",
                                     mde_curr ~ "Current vs. past MDD",
                                     NMR ~ "Nicotine metabolism ratio",
                                     Only.Menthol ~ "Exclusive mentholated cigarette user",
                                     readiness ~ "Readiness to quit smoking"),
              type = list(readiness ~ "continuous"),
              statistic = all_continuous() ~ "{mean} ({sd})",
              missing = "ifany",
              missing_text = "Missing") %>%
  add_overall(last = TRUE) %>%
  modify_spanning_header(update = all_stat_cols() ~ "**Behavioral and Pharmacological Treatment Assignment**") %>%
  modify_footnote(update = all_stat_cols() ~ "Mean (SD) for continuous; n (%) for categorical") %>%
  bold_labels()

summary_table %>%
  as_kable_extra(booktabs = TRUE, caption = "Participant Characteristics by Treatment Arm",
                 longtable = TRUE, linesep = "") %>%
  kableExtra::kable_styling(font_size = 7,
                            latex_options = c("repeat_header", "HOLD_position", "scale_down"))%>%
  column_spec(1, width = "3.5cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "2cm") %>%
  column_spec(5, width = "2cm") %>%
  column_spec(6, width = "2cm") %>%
  row_spec(0, bold = TRUE, font_size = 7)
```


To further investigate the variation among groups, we generate a series of stacked bar plots to illustrate the proportions of several key categorical variables by abstinence status and treatment arms, along with distribution plots for selected continuous variables. We can examine potential differences in these baseline characteristics across treatment and abstinence outcomes, understanding how these factors would influence or interact with our treatment effectiveness.

```{r, out.width = "100%", fig.pos = "H", fig.align = "center", fig.height = 4}
a <- ggplot(averaged_data_factor, aes(x = abst, fill = inc)) +
  geom_bar(position = "fill") +
  facet_wrap(~ trt) +
  labs(x = "Abstinence Status",
       y = "Proportion",
       fill = "Income Level") +
  theme_minimal() +
  scale_fill_brewer(palette = "GnBu") +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6)) +
  guides(fill = guide_legend(nrow = 3))

b <- ggplot(averaged_data_factor, aes(x = abst, fill = edu)) +
  geom_bar(position = "fill") +
  facet_wrap(~ trt) +
  labs(x = "Abstinence Status",
       y = "Proportion",
       fill = "Education Level") +
  theme_minimal() +
  scale_fill_brewer(palette = "GnBu") +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6)) +
  guides(fill = guide_legend(nrow = 2))

combined_plot <- (wrap_elements(panel = a + theme(legend.position = "bottom")) /
                  wrap_elements(panel = b + theme(legend.position = "bottom"))) +
  plot_layout(ncol = 2, guides = 'collect') +
  plot_annotation(title = "Figure 1: Baseline Characteristics by Abstinence Status and Treatment Group",
                  theme = theme(plot.title = element_text(size = 8, hjust = 0.5))) 

combined_plot <- combined_plot & theme(plot.margin = margin(10, 10, 10, 10),
                                       legend.position = c(0.5, 0.1))

combined_plot

ggplot(averaged_data_factor, aes(x = abst, fill = race)) +
  geom_bar(position = "fill") +
  facet_wrap(~ trt) +
  labs(title = "Figure 2: Race by Abstinence Status and Treatment Group",
       x = "Abstinence Status",
       y = "Proportion",
       fill = "Race Group") +
  theme_minimal() +
  scale_fill_brewer(palette = "GnBu") +
  theme(axis.title = element_text(size = 8),
        title = element_text(size = 10),
        axis.text = element_text(size = 8),
        legend.text = element_text(size = 6))
```


```{r, out.width = "100%", fig.pos = "H", fig.align = "center", fig.height = 4}
ftcd_score_plot <- ggplot(averaged_data_factor, aes(x = ftcd_score, fill = abst)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ trt) +
  labs(title = "FTCD Score",
       x = "FTCD Score",
       y = "Density",
       fill = "Abstinence Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "#FF9999", "Yes" = "#99CCFF")) +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6))

shaps_score_pq1_plot <- ggplot(averaged_data_factor, aes(x = shaps_score_pq1, fill = abst)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ trt) +
  labs(title = "Shaps Score",
       x = "Anhedonia",
       y = "Density",
       fill = "Abstinence Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "#FF9999", "Yes" = "#99CCFF")) +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6))

combined_plot2 <- (wrap_elements(panel = ftcd_score_plot + theme(legend.position = "bottom")) /
                  wrap_elements(panel = shaps_score_pq1_plot + theme(legend.position = "bottom"))) +
  plot_layout(ncol = 2, guides = 'collect') +
  plot_annotation(title = "Figure 1: Baseline Characteristics by Abstinence Status and Treatment Group",
                  theme = theme(plot.title = element_text(size = 8, hjust = 0.5))) 

combined_plot2 <- combined_plot2 & theme(plot.margin = margin(10, 10, 10, 10),
                                       legend.position = c(0.5, 0.1))

combined_plot2
```

```{r, out.width = "100%", fig.pos = "H", fig.align = "center", fig.height = 4}
readiness_plot <- ggplot(averaged_data_factor, aes(x = readiness, fill = abst)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ trt) +
  labs(title = "Readiness to Quit",
       x = "Readiness to Quit",
       y = "Density",
       fill = "Abstinence Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "#FF9999", "Yes" = "#99CCFF"))  +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6))

NMR_plot <- ggplot(averaged_data_factor, aes(x = NMR, fill = abst)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ trt) +
  labs(title = "NMR",
       x = "NMR",
       y = "Density",
       fill = "Abstinence Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "#FF9999", "Yes" = "#99CCFF"))  +
  theme(axis.title = element_text(size = 6),
        title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        strip.text = element_text(size = 6))

combined_plot3 <- (wrap_elements(panel = readiness_plot + theme(legend.position = "bottom")) /
                  wrap_elements(panel = NMR_plot + theme(legend.position = "bottom"))) +
  plot_layout(ncol = 2, guides = 'collect') +
  plot_annotation(title = "Figure : Baseline Characteristics by Abstinence Status and Treatment Group",
                  theme = theme(plot.title = element_text(size = 8, hjust = 0.5))) 

combined_plot3 <- combined_plot3 & theme(plot.margin = margin(10, 10, 10, 10),
                                       legend.position = c(0.5, 0.1))
combined_plot3
```

```{r, out.width = "80%", fig.pos = "H", fig.align = "center"}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 4, 0))
hist(averaged_data_factor$hedonsum_n_pq1, main = "", xlab = "hedonsum_n_pq1", 
     col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
hist(averaged_data_factor$hedonsum_y_pq1, main = "", xlab = "hedonsum_y_pq1", 
     col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
hist(averaged_data_factor$shaps_score_pq1, main = "", xlab = "shaps_score_pq1", 
     col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
hist(averaged_data_factor$NMR, main = "", xlab = "NMR", 
     col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
mtext("Figure : Distribution of Skewed Variables (Before Transformation)", outer = TRUE, cex = 0.8, font = 1)


# par(mfrow = c(2, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 4, 0))
# hist(averaged_data_factor$hedonsum_n_pq1, main = "", xlab = "hedonsum_n_pq1", 
#      col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
# hist(averaged_data_factor$hedonsum_y_pq1, main = "", xlab = "hedonsum_y_pq1", 
#      col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
# hist(averaged_data_factor$shaps_score_pq1, main = "", xlab = "shaps_score_pq1", 
#      col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
# hist(averaged_data_factor$NMR, main = "", xlab = "NMR", 
#      col = "lightblue", breaks = 15, cex.main = 1, cex.lab = 1, cex.axis = 0.8, font.main = 1)
# mtext("Figure : Distribution of Skewed Variables (After Transformation)", outer = TRUE, cex = 0.8, font = 1)
```

```{r}
# multiple imputation with m = 5
imputed_data <- mice(data, m = 5, method = 'pmm', maxit = 50, seed = 2550, printFlag = FALSE)

# extract the five imputed datasets
completed_datasets <- list()
for (i in 1:5) {
  completed_datasets[[i]] <- complete(imputed_data, i)
}

# calculate average/mode of each missing variable
averaged_data <- completed_datasets[[1]]

for (var in names(averaged_data)) {
  if (any(is.na(data[[var]]))) {
    if (is.numeric(averaged_data[[var]])) {
      averaged_data[[var]] <- rowMeans(sapply(completed_datasets, function(x) x[[var]]))
    } else {
      averaged_data[[var]] <- apply(sapply(completed_datasets, function(x) x[[var]]), 1, function(vals) {
        vals <- as.factor(vals)
        unique_vals <- unique(vals)
        unique_vals[which.max(tabulate(match(vals, unique_vals)))]
      })
    }
  }
}
```

```{r}
new_data <- averaged_data %>%
  mutate(race = as.factor(case_when(Black == 0 & Hisp == 0 & NHW == 0 ~ "Unknown",
                                    Black == 1 & Hisp == 1 & NHW == 1 ~ "Mixed Race",
                                    Black == 1 & Hisp == 1 ~ "Mixed Race",
                                    Black == 1 & NHW == 1 ~ "Mixed Race",
                                    NHW == 1 & Hisp == 1 ~ "Mixed Race",
                                    Black == 1 ~ "Black",
                                    Hisp == 1 ~ "Hispanic",
                                    NHW == 1 ~ "Non-Hispanic White",
                                    TRUE ~ "Other")),
         trt = as.factor(case_when(Var == 1 & BA == 1 ~ "BASC + varenicline",
                         Var == 0 & BA == 1 ~ "BASC + placebo",
                         Var == 1 & BA == 0 ~ "ST + varenicline",
                         Var == 0 & BA == 0 ~ "ST + placebo",
                         TRUE ~ NA_character_)),
         inc = fct_recode(as.factor(inc), 
                          "Less than $20,000" = "1", 
                          "$20,000-35,000" = "2", 
                          "$35,001-50,000" = "3", 
                          "$50,001-75,000" = "4", 
                          "More than $75,000" = "5"),
         edu = fct_recode(as.factor(edu), 
                          "Grade School" = "1", 
                          "Some high school" = "2", 
                          "High school graduate or GED" = "3", 
                          "Some college/technical school" = "4", 
                          "College graduate" = "5"))

new_data <- new_data %>%
  mutate(inc = fct_relevel(inc, "Less than $20,000", "$20,000-35,000", 
                           "$35,001-50,000", "$50,001-75,000", "More than $75,000"),
         edu = fct_relevel(edu, "Grade School", "Some high school", "High school graduate or GED",
                           "Some college/technical school", "College graduate"))

# new_data$hedonsum_n_pq1 <- log(new_data$hedonsum_n_pq1)
# new_data$hedonsum_y_pq1 <- log(new_data$hedonsum_y_pq1)
# new_data$shaps_score_pq1 <- log(new_data$shaps_score_pq1)
# new_data$NMR <- log(new_data$NMR)
```

```{r}
set.seed(2550)
train_index <- createDataPartition(new_data$trt, p = 0.7, list = FALSE)
train_data <- new_data[train_index, ]
test_data <- new_data[-train_index, ]

X <- model.matrix(abst ~ trt * (age_ps + sex_ps + inc + edu + ftcd_score + ftcd.5.mins +
                  bdi_score_w00 + cpd_ps + crv_total_pq1 + hedonsum_n_pq1 + hedonsum_y_pq1 +
                  shaps_score_pq1 + otherdiag + antidepmed + mde_curr + NMR + Only.Menthol +
                  readiness + race), data = train_data)[, -1]
y <- train_data$abst
cv_model <- cv.glmnet(X, y, family = "binomial", alpha = 1, nfolds = 10, nlambda = 100)
plot(cv_model)
best_lambda <- cv_model$lambda.min
lasso_model <- glmnet(X, y, family = "binomial", alpha = 1, lambda = best_lambda)

coefficients <- as.data.frame(as.matrix(coef(lasso_model)))
coefficients$variable_name <- rownames(coefficients)
rownames(coefficients) <- NULL
colnames(coefficients) <- c("Estimates", "Variable")
coefficients <- coefficients[, c("Variable", "Estimates", setdiff(names(coefficients), c("Estimates", "Variable")))] %>%
  filter(Estimates != 0)
```

```{r}
prediction_train <- predict(lasso_model, X, type = "response")
roc_train <- roc(y, prediction_train)
auc_train <- auc(roc_train)
print(auc_train)
```

```{r}
# X_interaction <- model.matrix(abst ~ trt * (ftcd_score + NMR + race + mde_curr), 
#                               data = train_data)[, -1]
# y_interaction <- train_data$abst
# cv_model_interaction <- cv.glmnet(X_interaction, y_interaction, family = "binomial", 
#                                   alpha = 1, nfolds = 10, nlambda = 100) 
# best_lambda_interaction <- cv_model_interaction$lambda.min
# lasso_model_interaction <- glmnet(X_interaction, y_interaction, family = "binomial", 
#                                   alpha = 1, lambda = best_lambda_interaction)
# 
# prediction_train <- predict(lasso_model_interaction, X_interaction, type = "response")
# roc_train <- roc(y_interaction, prediction_train)
# auc_train <- auc(roc_train)
# print(auc_train)
```

```{r}
X_test_interaction <- model.matrix(abst ~ trt * (age_ps + sex_ps + inc + edu + ftcd_score + ftcd.5.mins +
                  bdi_score_w00 + cpd_ps + crv_total_pq1 + hedonsum_n_pq1 + hedonsum_y_pq1 +
                  shaps_score_pq1 + otherdiag + antidepmed + mde_curr + NMR + Only.Menthol +
                  readiness + race), data = test_data)[, -1]
y_test_interaction <- test_data$abst
prediction_test <- predict(lasso_model, X_test_interaction, type = "response")

roc_test <- roc(y_test_interaction, prediction_test)
auc_test <- auc(roc_test)
```

```{r, out.width = "80%", fig.pos = "H", fig.align = "center", fig.height = 4}
par(mfrow= c(1,2), oma = c(0, 0, 2, 0))
plot(roc_train, main = "Train Data", font.main = 1, cex.main = 0.8)
text(0.3, 0.2, paste("AUC =", round(auc_train, 3)), col = "blue", cex = 0.7)

plot(roc_test, main = "Test Data", font.main = 1, cex.main = 0.8)
text(0.3, 0.2, paste("AUC =", round(auc_test, 3)), col = "blue", cex = 0.7)
```

```{r, out.width = "80%", fig.pos = "H", fig.align = "center", fig.height = 4}
train_data$pred <- prediction_train
test_data$pred <- prediction_test
train_data <- train_data %>%
  mutate(abst_num = as.numeric(as.character(abst)))
test_data <- test_data %>%
  mutate(abst_num = as.numeric(as.character(abst)))
cal_plot_train <- calibration_plot(data = train_data, obs = "abst_num", pred = "pred", title = "Train Data", y_lim = c(0, 1), x_lim=c(0, 1))
cal_plot_test <- calibration_plot(data = test_data, obs = "abst_num", pred = "pred", title = "Test Data", y_lim = c(0, 1), x_lim=c(0, 1))

grid.arrange(cal_plot_train$calibration_plot, 
             cal_plot_test$calibration_plot, ncol = 2,
             top = text_grob("Figure 4: Calibration Plot Comparison"))
```
\newpage
# Appendix

```{r ref.label = knitr::all_labels()}
#| echo: true
#| eval: false
```